---
subtitle: "TMA4268 Statistical Learning V2024"
title: "Compulsory exercise 1: Group 49"
author: "Jasper Steinberg, Maximilian Rønseth"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  # html_document
  pdf_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE,tidy=TRUE,message=FALSE,warning=FALSE,strip.white=TRUE,prompt=FALSE,
                      cache=TRUE, size="scriptsize",fig.width=4, fig.height=3,fig.align = "center")

```

```{r,eval=TRUE,echo=FALSE}
#install.packages("knitr") # probably already installed
#install.packages("rmarkdown") # probably already installed
#install.packages("ggplot2") # plotting with ggplot2
#install.packages("dplyr") # for data cleaning and preparation
#install.packages("ggfortify") # for model checking
#install.packages("MASS")
#install.packages("tidyr")
#install.packages("carData") # dataset
#install.packages("class")
#install.packages("pROC")
#install.packages("plotROC")
#install.packages("boot")
#install.packages("ggmosaic")
library("knitr")
library("rmarkdown")
library("ggplot2")
library("dplyr")
library("ggfortify")
library("boot")
```

<!--  Etc (load all packages needed). -->

# Problem 1

## a)

Examples of qualitative variables are eye colour (green, blue, brown),
coffee roast (dark, medium, light) and sex (male, female).

Examples of quantitative variables are height (0-300cm), light (in
lumen) and speed (in m/s).

## b)

KNN can be used, since it is simply a method on categorizing data based
on the category of its neighbors. Linear regression can not be used as
it produces probabilities outside of $[0,1]$. Logistic regression can be
used as it is always bounded between $[0,1]$. LDA can be used, since it
is fundamentaly about prediction given the data conditioned on the
response. Since QDA only differs from LDA in assumptions, it is also
applicable.

## c)

### i)

The term $E[f(X) - \hat{f}(X)]^2$ corresponds to the bias of the model,
i.e. the error between the real data and the model. The model variance
is given by the term $\text{Var}(\hat{f}(X))$, i.e. how much the model
varies when we change the data. The last term $\text{Var}(\epsilon)$
represents the irreducible error, omnipresent in all of science.

### ii)

The bias-variance trade-off is that since the LHS is a fixed number and
the variance of the irreducible error always exists, changing the bias
will lead to change in the variance and vice versa (to maintain equality
in the expected mean squared error expression).

### iii)

Since the variance is given by $\text{Var}(x) = E(x^2) - E(x)^2$, we
have that
$$E[(Y - \hat{Y})^2] = \text{Var}(Y - \hat{Y}) + E[(Y-\hat{Y})]^2.$$ Now
we insert the expressions in terms of $f$ to obtain
$$= \text{Var}(f(X) + \epsilon - \hat{f}(X)) + E(f(X) - \hat{f}(X) + \epsilon)^2.$$
First handling the variance term, we use that $f$ deterministic (i.e.
has $0$ variance). Thus, by also the standard property of the
variance,$\text{Var}(aX + Y) = a^2\text{Var}(X) + \text{Var}(Y)$, we
have
$$\text{Var}(f(X) + \epsilon - \hat{f}(X)) = \text{Var}(\hat{f}(X)) + \text{Var}(\epsilon).$$
To handle the expectation term, we use that $E$ is linear and that the
irreducible error has expectation $0$. Thus
$$E(f(X) - \hat{f}(X) + \epsilon)^2 = E(f(X) - \hat{f}(X))^2.$$
Combining what we found we can conclude that the original expression
reduces to

$$=\text{Var}(\hat{f}(X)) + \text{Var}(\epsilon) + E(f(X) - \hat{f}(X))^2,$$
which proves the result.

## d)

We see that for $K=1$ the nearest neighbor is blue, thus the KNN
classification of the black dot would be blue. For $K=3$ the nearest
neighbors are blue, red, red, thus the black dot is classified as red.
For $K=5$ we count 2 blue and 3 red dots, thus the classification yields
red.

## e)

### i)

```{r,echo=TRUE,eval=TRUE}

library(MASS) 
#data(Boston)

lm1 <- lm(medv ~ rm + age, data = Boston)

summary(lm1)
```

### ii)

```{r,echo=TRUE,eval=TRUE}
indexLM <- c(6, 7, 14) 

reducedBoston <- Boston[, indexLM]

cor_matrix <- cor(reducedBoston) 

cor_matrix
```

### iii)

```{r,echo=TRUE,eval=TRUE}

lm2 <- lm(medv ~ rm + age + nox, data = Boston) 
summary(lm2)

```

### iv)

The change could attributed to compounding effects, essentially saying
that the age and amount of air pollution are correlated variables. Below
we compute the correlation and see that the variables are highly
correlated.

```{r,echo=TRUE,eval=TRUE}
cor(Boston)[5,7]
```

# Problem 2

## a)

### i)

```{r, eval=FALSE, echo=TRUE}
# load the boston housing price dataset
data(Boston)


# fit the linear regression model
lm_model <- lm(medv ~ crim*age + rm + I(rm^2), Boston)

summary(lm_model)

```

### ii)

We first note that all terms not involving $X_{\text{crim}}$ will
cancel, since they remain the same. Thus the expression for the change in housing prices will read
$$\Delta Y_{\text{medv}} = \beta_1 \Delta X_{\text{crim}} + \beta_3 \Delta X_{\text{crim}}X_{\text{age}}.$$ Given that crime is reduced by $10$, we have that the change in crime will be $\Delta X_{\text{crim}} = -10$. Furthermore, we were given that $X_{\text{age}} = 60$. To conclude we only need the coefficients, which we extract from the previous task as $\beta_1 = -0.796544$ and $\beta_3 = 0.005792$. Inserting all the numbers we find
$\Delta Y_{\text{medv}}\simeq 4.49.$

## b)

If we want to reduce uncertainty in the model parameters we could
increase the sample size, since this would decrease the standard error
of the parameter estimates.

## c)

### i)

If we were to use $\hat{\beta}_3 = 10$ in the formula for the t-value,
we get
$$t = \frac{\hat{\beta}_3}{\text{SE}(\hat{\beta}_3)} = \frac{10}{0.40201} = 24.875.$$

###ii)

Yes, we see that the F-statistic ($=216.1$) in the summary has an
extremely low p-value, $p < 2.2e-16$, thus we expect at least one of the
predictors to be helpful in predicting the response.

### iii)

```{r,echo=TRUE,eval=TRUE}
lm_model2 <- lm(medv ~ crim + age, data = Boston) 
summary(lm_model2)
```

Thus, even with this reduced model, we would still find it highly
probable that at least one of the predictors would be beneficial to
predicting the response; this is due to the extremely low $p$-value
associated with the F-test.

## d)

```{r,echo=TRUE,eval=TRUE}

#We override the previous lm_model to fit the task
lm_model <- lm(medv ~ crim + age + rm, data = Boston)

```

### i)

```{r,echo=TRUE,eval=TRUE}

new_data <- data.frame(crim = 10, age = 90, rm = 5)

conf_int <- predict(lm_model, newdata = new_data, interval =
"confidence", level = 0.99)

#Lower bound 
conf_int[1, 2]

#Upper bound 
conf_int[1, 3]
```

### ii)

```{r,echo=TRUE,eval=TRUE}

pred_int <- predict(lm_model, newdata = new_data, interval =
"prediction", level = 0.99)

#Lower bound 
pred_int[1, 2]

#Upper bound 
pred_int[1, 3]
```

### iii)

The confidence interval of level 99% is defined in the following way:
given some parameter you are trying to estimate, we will in 99% of times
find that the true value of the parameter lies in our interval.

A prediction interval, is defined in the following: if you sample new
data, then in 99% of the cases , we would expect our new value to lie
within this interval.

### iv)

```{r,echo=TRUE,eval=TRUE}


autoplot(lm_model, smooth.colour = NA)

```

The Tukey-Anscombe plot has a slight curvature, in addition to outlier
points on a line, indicating that our linear modeling assumption is
problematic. The QQ-plot shows that the normal assumption is violated,
as it deviates sharply from the linear line towards the end.

## e)

### i)

The problem with the model is that is is not identifiable. In other
words we could add an arbitrary constant to $\beta_0$, and subtract it
from $\beta_1$ and $\beta_2$ without changing the model. Thus we loose
the ability to identify the optimal parameters.

### ii)

To solve the aforementioned problem we assign one of the groups as the
reference group. The correct model would then read

$$
y = 
\begin{cases}
\beta_0 + \beta_1 + \epsilon  \text{  , if  } x_{male} = 1
\\
\beta_0 + \mathbb{\epsilon} \text{      ,          if } x_{male} = 0
\end{cases}
$$

### iii)

If we first define $\text{Bachelor} = 1$, $\text{Master} = 2$ and
$\text{PhD} = 3$. We also set the reference variable as $Bachelor$, i.e.
$\beta_1 = 0$. Thus the model would read $$ y =
\begin{cases}
\beta_0 + \epsilon  \text{,           if  } x_{1} = 1
\\
\beta_0 + \beta_2 + \mathbb{\epsilon} \text{,  if } x_{2} = 1
\\
\beta_0 + \beta_3 + \epsilon \text{,  if } x_{3} = 1
\end{cases}
$$

## f)

i)  TRUE ii) FALSE iii) TRUE iv) FALSE

# Problem 3

## a)

### i)

```{r,echo=TRUE,eval=TRUE}

set.seed(123)

# prepare the dataset into training and test datasets
#install.packages("titanic")
library(titanic)

data("titanic_train")

# remove some variables that are difficult to handle.
# NB! after the removal, the datasets have the variable names of
# [Survived, Pclass, Sex, Age, SibSp, Parch, Fare].
vars_to_be_removed <- c("PassengerId", "Name", "Ticket", "Cabin", "Embarked")
titanic_train <- titanic_train[, -which(names(titanic_train) %in% vars_to_be_removed)]

# make Pclass a categorical variable
titanic_train$Pclass <- as.factor(titanic_train$Pclass)

# divide the dataset into training and test datasets
train_idx <- sample(1:nrow(titanic_train), 0.8 * nrow(titanic_train))
titanic_test <- titanic_train[-train_idx, ]
titanic_train <- titanic_train[train_idx, ]

# remove the rows with missing values
titanic_train <- na.omit(titanic_train)
titanic_test <- na.omit(titanic_test)

# [ TODO] fit the logistic regression model
logReg <- glm(Survived ~ ., data = titanic_train, family = binomial) 
summary(logReg)

# [ TODO] compute the accuracy on the test set

# Make predictions on test set
predicted_probs <- predict(logReg, newdata = titanic_test, type = "response")


predicted_class <- ifelse(predicted_probs > 0.5, 1, 0)

# Calculate accuracy
accuracy <- mean(predicted_class == titanic_test$Survived)

accuracy
```

### ii)

```{r,echo=TRUE,eval=TRUE}
anova(logReg, test = "Chisq")
```

We see that the p-value for Pclass is extremely low, less than machine
epsilon. Thus Pclass seems like a very relevant predictor for survival.

### iii)

```{r,echo=TRUE,eval=TRUE}

c_data <- data.frame(Pclass = factor(c(1, 3)), Sex = "female", Age = 40, SibSp = c(1, 1), Parch = c(0,0), Fare = c(200, 20))

predict(logReg, newdata = c_data, type = "response")
  
```

### iv)

```{r,echo=TRUE,eval=TRUE}

LDA_Reg <- lda(Survived ~ ., data = titanic_train) 
LDA_Reg

predicted_probs2 <- predict(LDA_Reg, newdata = titanic_test, type = "response")


# Calculate accuracy
mean(predicted_probs2$class == titanic_test$Survived)

```

### iv)

```{r,echo=TRUE,eval=TRUE}
QDA_Reg <- qda(Survived ~ ., data = titanic_train) 
QDA_Reg

predicted_probs3 <- predict(QDA_Reg, newdata = titanic_test, type = "response")


# Calculate accuracy
mean(predicted_probs3$class == titanic_test$Survived)

```

### vi)

```{r,echo=TRUE,eval=TRUE}
library(pROC)

roc_logReg <- roc(titanic_test$Survived, predicted_probs)
roc_lda <- roc(titanic_test$Survived, predicted_probs2$posterior[, "1"])
roc_qda <- roc(titanic_test$Survived, predicted_probs3$posterior[, "1"])


```

```{r,echo=TRUE,eval=TRUE}
plot(roc_logReg, main = "Roc plot for logReg")
plot(roc_lda, main = "Roc plot for LDA")
plot(roc_qda, main = "Roc plot for QDA")
```

### vii)

```{r,echo=TRUE,eval=TRUE}
auc(roc_logReg)
auc(roc_lda)
auc(roc_qda)
```

### viii)

Based on the AUC-scores, we see that the closes to $1$, and thus the
best was logistic regression. The worst was QDA. We note that they are
very close numerically, thus it is hard to draw solid conclusions based
on this test.

## b)

### i)

In essence the methodologies differ, in the sense that the diagnostic
paradigm we directly infer on $P(Y = i\mid X = x)$, where $i$ is some
group classification, and $x$ is the data. The sampling paradigm instead
does inference on the priors and reverse condition through Bayes' rule.

### ii)

Logistic regression, Naive Bayes classifier, LDA, and QDA belong to the
diagnostic paradigm, while KNN belongs to the sampling paradigm.

## c)

### i)

Using the notation in the textbook, we can find the decision boundary by
solving $\delta_1(x) = \delta_2(x)$. In our case $\mu_1 = -2$,
$\mu_2 = 2$ and $\sigma_1 = \sigma_2 = 1.5^2$, thus we solve $$
x\frac{\mu_1}{\sigma_1^2} - \frac{\mu_1^2}{2\sigma_1^2} + \log(\pi_1)
= x\frac{\mu_2}{\sigma_1^2} - \frac{\mu_2^2}{2\sigma_1^2} + \log(\pi_2)$$
Inserting numbers and doing the calculating we find
$-\frac{9}{16}\log\left(\frac{7}{3}\right) \simeq -0.47665$.

### ii)

```{r,echo=TRUE,eval=TRUE}
set.seed(123) # Replace 123 with any number of your choice

# generate data for the two normal distributions
n_samples_class1 <- 3000
n_samples_class2 <- 7000

x1 <- rnorm(n_samples_class1, mean = -2, sd = 1.5)
x2 <- rnorm(n_samples_class2, mean = 2, sd = 1.5)

# create a data frame with the generated data
df <- data.frame(X = c(x1, x2), class = c(rep(1, n_samples_class1), rep(2, n_samples_class2)))
#summary(df)

lda_model <- lda(class ~ X, data = df)

```

### iii)

```{r,echo=TRUE,eval=TRUE}

pred_post <- predict(lda_model, newdata = df, type = "response")

p_1_x <- pred_post$posterior[, "1"]

p_2_x <- pred_post$posterior[, "2"]

```

### iv)

```{r,echo=TRUE,eval=TRUE}
plot(df$X, p_1_x, col = "blue", xlab = "X", ylab = "Probability", main = "Posterior Probabilities")
points(df$X, p_2_x, col = "red")
legend("topright", legend = c("p1(X)", "p2(X)"), col = c("blue", "red"), lty = 1)

```

## d)

i)  TRUE ii) TRUE iii) TRUE iv) TRUE

# Problem 4

## a)

iv) 

## b)

### i)

```{r,echo=TRUE,eval=TRUE}

set.seed(123)

# Import the Boston housing price dataset

library(MASS)
library(caret)
data(Boston)

# select specific variables
selected_vars <- c("crim", "rm", "age", "medv")
boston_selected <- Boston[, selected_vars]

# manually perform the 5-fold cross-validation


#folds <- createFolds(boston_selected$medv, k = 4)
#Corrected: k=5
folds <- createFolds(boston_selected$medv, k = 5)

#Size of the data, to do LOOCV set k = 506
length(boston_selected$medv)

rmse_list <- list()
for (i in 1:length(folds)) {
  # get the training and validation sets
  
  #Change the minus
  train <- boston_selected[-folds[[i]], ]
  val <- boston_selected[folds[[i]], ]
  
  # fit a linear regression model
  model <- lm(medv ~ ., data = train)
  
  # compute RMSE on the validation set
  pred <- predict(model, val)
  
  #rmse <- sqrt(mean((pred - val$medv))) # root mean squared error (RSME)
  #Corrected: Should be squared error
  rmse <- sqrt(mean((pred - val$medv)^2))
  
  rmse <- rmse[1] # take out the value
  
  # store rmse in rmse_list
  #rmse_list[[i]] <- rmse
  #Corrected: Object is a list
  rmse_list[i] <- rmse
}

# compute mean of rmse_list
rmse_mean <- mean(as.numeric(rmse_list))

cat("rmse_mean:", rmse_mean, "\n")

```

### ii)

To use LOOCV instead, we change $k=5$ into $k=506$, since its the size of he data. We commented in the code above.

## c)

### i)

```{r,echo=TRUE,eval=TRUE}

# simulate data (no need to change this part)
set.seed(123)
n <- 1000 # population size
dataset <- rnorm(n) # population

# bootstrap

#In the context of the task, we accept B=10, since the population size is quite low.
B <- 10 # bootstrap sample size

#There is no need to make a matrix in this case, we change to a vector. 
boot <- rep(NA, B)

for (i in 1:B) {
  
#We should sample the size of the dataset, not 1. Also bootstraping uses replacement. 
boot[i] <- median(sample(dataset, n, replace = TRUE))
}

# compute the standard error of the median from the bootstrap samples
#We correct grammar and give a more readable name.
standard_error <- sd(boot)
cat("Estimated standard error:", standard_error, "\n")

```

### ii)

```{r,echo=TRUE,eval=TRUE}

# simulate data (no need to change this part)
set.seed(123)
n <- 1000 # population size
dataset <- rnorm(n) # population
# bootstrap
B <- 10 # bootstrap sample size
boot <- matrix(NA, nrow = B, ncol = 1)
for (i in 1:B) {
  boot[i, ] <- median(sample(dataset, 1, replace = FALSE))
}
# compute the standard error of the median from the bootstrap samples
standard_erorr_of_the_median_bootstrap <- sd(boot)
cat("standard_erorr_of_the_median_bootstrap:", standard_erorr_of_the_median_bootstrap, "\n")

```

Without the corrections, we just find the estimated variance of the
sampling distribution i.e. a number close to $1$. In the corrected
version, we find the deviation of the expectation, i.e. a number close
to $0$.

## d)

i)  FALSE ii) TRUE iii) FALSE iv) TRUE
